A Mechanical Formalization of Higher-Ranked Polymorphic Type Inference

### 2019-02-11 sections/introduction.tex
Intro WIP

Modern functional programming languages, such as Haskell or OCaml,
use sophisticated forms of type inference. The type systems of these languages are
descendents of Hindley-Milner~\cite{}, which was revolutionary at the
time in allowing type-inference to proceed without any type
annotation. The traditional Hindley-Milner type system supports
top-level \emph{implicit (parametric) polymorphism}. With implicit
polymorphism, type arguments of polymorphic functions are
automatically instantiated. Thus the Hindley-Milner type system 
strikes a great balance between expressive power and usability.

As functional languages evolved the need for more expressive
power has motivated language designers to look beyond Hindley-Milner.
In particular one popular direction has been to allow 
\emph{higher-ranked polymorphism} where polymorphic types can
occur anywhere in a type signature.  
An important challenge is that full type inference for higher-ranked
polymorphism is know to be undecidable~\cite{}. Therefore some type
annotations are necessary to guide type inference. In response to this 
challenge several decidable type systems requiring some annotations 
have been proposed~\cite{}. Two closely related type systems that 
support \emph{predicative} higher-ranked type inference were proposed 
by Peyton Jones et al.~\cite{} and Dunfield and
Krishnaswami (henceforth denoted as DK). 
These type systems are
popular among language designer and their ideas have been adopted by
several modern functional languages, including Haskell and PureScript.
In those type systems
type annotations are required for polymorphic arguments of functions,
but other type annotations can be omitted. A canonical example is:
\bruno{the following text needs to be changed, since it is basically
  as in ITP.}
\begin{verbatim}
hpoly = (\f :: forall a. a -> a) -> (f 1, f 'c')
\end{verbatim}
The function \verb|hpoly| cannot be
type-checked in Hindley-Milner.  The type of \verb|hpoly| is the rank-2 type:
\verb|(forall a. a -> a) -> (Int, Char)|. Notably (and unlike
Hindley-Milner) the lambda argument \verb|f| requires a
\emph{polymorphic} type annotation.
This annotation is needed because the single universal quantifier
does not appear at the top-level. Instead it is used to quantify a
type variable \verb|a| used in the first argument of the
function. 
Despite these additional annotations, Peyton Jones et al. and DK's
type inference algorithms preserve many of the desirable properties 
of Hindley-Milner. For example the applications of \verb|f| implicitly 
instantiate the polymorphic type arguments of \verb|f|.

@@

This paper presents the first full mechanical formalization of the metatheory for
higher-ranked polymorphic type inference. The system that we formalize
is the bidirectional type system by Dunfield and
Krishnaswami (DK). The DK type system has two variants (a declarative
and an algorithmic one) that have been been \emph{manually} proved to be
\emph{sound}, \emph{complete} and \emph{decidable}.

We present 
a mechanical formalization in the Abella theorem prover 
for DK's declarative type system using a different algorithm. We use a
new algorithm for a few different reasons. Firstly, our new algorithm
employs \emph{worklist judgements}, which allow precise capture of the
scope of vairables and simplify the formalization of scoping in a
theorem prover. Secondly, while DK's original formalization comes with very
well-written manual proofs, there are several details missing,
which complicate the task of writing a formla proof. 
Perhaps more importantly, some proofs/lemmas are wrong and it is
not clear how to replace these by other valid proofs/lemmas. 
\bruno{others}

Despite the use of a different algorithm we prove the
same results by DK, although with significantly different proofs and
proof techniques.

The calculus and its metatheory
have been fully formalized in the Abella theorem prover\cite{AbellaDesc}. % Add at least one ref to make compile work normally

@@

\item {\bf A full mechanical formalization of type inference with
    higher-ranked types:}

\item {\bf A new algorithm for DK:}

@@

\item {\bf Abella formalization:} The calculus and its metatheory
have been fully formalized in the Abella theorem prover\cite{AbellaDesc}.

>>>>

Modern functional programming languages, such as Haskell or OCaml,
use sophisticated forms of type inference. The type systems of these languages are
descendents of Hindley-Milner~\cite{}, which was revolutionary at the
time in allowing type-inference to proceed without any type
annotation. The traditional Hindley-Milner type system supports
top-level \emph{implicit (parametric) polymorphism}. With implicit
polymorphism, type arguments of polymorphic functions are
automatically instantiated. Thus implicit polymorphism and the absense of
type annotations mean that the Hindley-Milner type system 
strikes a great balance between expressive power and usability.

As functional languages evolved the need for more expressive
power has motivated language designers to look beyond Hindley-Milner.
In particular one popular direction has been to allow 
\emph{higher-ranked polymorphism} where polymorphic types can
occur anywhere in a type signature.  
An important challenge is that full type inference for higher-ranked
polymorphism is know to be undecidable~\cite{}. Therefore some type
annotations are necessary to guide type inference. In response to this 
challenge several decidable type systems requiring some annotations 
have been proposed~\cite{}. Two closely related type systems that 
support \emph{predicative} higher-ranked type inference were proposed 
by Peyton Jones et al.~\cite{} and Dunfield and
Krishnaswami (henceforth denoted as DK). 
These type systems are
popular among language designers and their ideas have been adopted by
several modern functional languages, including Haskell and PureScript.
In those type systems
type annotations are required for polymorphic arguments of functions,
but other type annotations can be omitted. A canonical example (here written in Haskell) is:
\begin{verbatim}
hpoly = (\f :: forall a. a -> a) -> (f 1, f 'c')
\end{verbatim}
The function \verb|hpoly| cannot be
type-checked in Hindley-Milner.  The type of \verb|hpoly| is the rank-2 type:
\verb|(forall a. a -> a) -> (Int, Char)|. Notably (and unlike
Hindley-Milner) the lambda argument \verb|f| requires a
\emph{polymorphic} type annotation.
This annotation is needed because the single universal quantifier
does not appear at the top-level. Instead it is used to quantify a
type variable \verb|a| used in the first argument of the
function. 
Despite these additional annotations, Peyton Jones et al. and DK's
type inference algorithms preserve many of the desirable properties 
of Hindley-Milner. For example the applications of \verb|f| implicitly 
instantiate the polymorphic type arguments of \verb|f|.

@@

This paper presents the first full mechanical formalization of the
metatheory for higher-ranked polymorphic type inference.
The system
that we formalize is the bidirectional type system by Dunfield and
Krishnaswami (DK)~\cite{}. We chose DK's type system because it is
quite elegant, well-documented and it comes with detailed manually
written proofs. Furthermore the system is adopted in practice by a few
real implementations of functional languages, including PureScript and
\bruno{others}.  The DK type system has two variants: a declarative
and an algorithmic one. The two variants have been been
\emph{manually} proved to be \emph{sound}, \emph{complete} and
\emph{decidable}.
We present a mechanical formalization in the Abella theorem prover for
DK's declarative type system using a different algorithm. While our
initial goal was to formalize both DK's declarative and algorithmic
versions, we faced technical challenges with the later, prompting us to find
an alternative.

One key challenge in type inference is variable binding,
because the algorithms typically do not rely simply on local
environments, but instead propagate information across judgements.
Yet, there is little work on how to deal with these complex forms of
binding in theorem provers. To keep track of variable scoping, DK's
algorithmic version employs input and output contexts to track
information that is discovered throught type inference. However
modelling output contexts in a theorem prover is non-trivial. Our
work takes a different approach by refining the idea of
\emph{worklists judgments}~\cite{}, proposed recently to mechanically
formalize an algorithm for \emph{polymorphic subtyping}.
However, a key challenge not addressed previously with
worklist judgements was how to deal with inference judgements.
In our work we show how worklist judgements can be extended to
deal with the inference judgements that are necessary for type-inference.
Furthermore we also present new techniques that enable precise tracking
of the scope of variables.

%% Polymorphic subtyping is one component of higher-ranked type inference, and
%% it is also a key part of DK's system.

Another challenge that we faced were missing details and sometimes
incorrect proofs/lemmas in DK's formalization. While DK's original
formalization comes with very well-written manual proofs, there are
several details missing, which complicate the task of writing a
mechanically verified proof. Perhaps more importantly, some
proofs/lemmas are wrong and it is not clear how to replace these by
other valid proofs/lemmas.

Despite the use of a different algorithm we prove the
same results by DK, although with significantly different proofs and
proof techniques. The calculus and its metatheory
have been fully formalized in the Abella theorem prover~\cite{AbellaDesc}. % Add at least one ref to make compile work normally

@@

\item {\bf A full mechanical formalization of type inference with
  higher-ranked types:} Our work presents the first full mechanical formalization
  for type inference of higher ranked types. The formalization is done in the
  Abella theorem prover\cite{AbellaDesc} and it is available
  online\footnote{{\{bf Note to reviewers:} Due to the anonymous submission process
  the url is sent as part of the supplementary materials.}

\item {\bf A new algorithm for DK:} Our work proposes a novel algorithm that implements
  DK's declaratiuve bidirectional type systems. We prove the soundness ...

@@

\item {\bf Abella formalization:} The calculus and its metatheory
have been fully formalized in the .
